---
title: "supreme_court"
format: html
editor: visual
---

# Supreme Court Polarization

### Goal: To explore whether collaboration is decreasing within the Court due to partisanship. 

```{r}
library(tidyverse)
library(ggiraph)
library(patchwork)
```

Getting Data from the Supreme Court Database at Washington University Law

```{r}

zip_url <- "http://scdb.wustl.edu/_brickFiles/2024_01/SCDB_2024_01_justiceCentered_Citation.Rdata.zip"

zip_destfile <- "SCDB_2024_01_justiceCentered_Citation.Rdata.zip"

download.file(zip_url, zip_destfile, mode = "wb")

unzip(zip_destfile, exdir = ".")

load("SCDB_2024_01_justiceCentered_Citation.Rdata")

```

I might try this without knowledge of their party affiliations AND independently of it as well.

### Q1: How do I track who votes together? 

What are the groupings? Find votes where they vote the same way, and see how often they appear on the same side together.

### Q2: Is the mean concurring vote total changing over years? 6-3

```{r}
##making a subset
cases_selected <- SCDB_2024_01_justiceCentered_Citation |>
  select(caseId,
         dateDecision,
         majVotes,
         minVotes,
         justice,
         justiceName,
         vote,
         direction,
         majority) |>
  mutate( year = year(dateDecision),
          agreement_percentage = if_else(
            minVotes == 0,
            100,  # Set to 100% if unanimous
            (majVotes / (majVotes + minVotes)) * 100 
            )
          )


```

```{r}
cases_selected |> 
  group_by(year)|>
  summarize(avg_agree_perc = mean(agreement_percentage))|>
  ggplot(aes(x=year, y = avg_agree_perc))+
  geom_point()+
  geom_line()+
  geom_smooth(method = "loess", span = 0.5, 
              method.args = list(degree = 2, #linear or quadratic
                                 family = "symmetric"), #symmetric or gaussian
              na.rm = TRUE)+  
  labs(title = "Supreme Court Agreement Percentage")+
  theme_minimal()

```

Conclusion: There has been a slight upward trajectory to greater agreement in the court. This could just represent partisan sorting or partisan majorities in the modern era. What I need to measure is ideological sorting. Has that gotten worse or better?

### Q3 Does the direction variable actually track changes in the court?

```{r}
cases_selected |>
  group_by(year)|>
  summarize(avg_direction = mean(direction, na.rm = TRUE)) |>
  ggplot(aes(x = year, y = avg_direction, color = avg_direction)) +
  geom_line(size = 1.25)+
  geom_point()+
  scale_color_gradient2(
    low = "red",    # Color for low values
    mid = "gray",   # Color for midpoint
    high = "darkblue",    # Color for high values
    midpoint = 1.5     # Set the midpoint value for the diverging scale
  ) +
  geom_smooth(method = "loess", span = 0.5, 
              method.args = list(degree = 2, #linear or quadratic
                                 family = "symmetric"), #symmetric or gaussian
              na.rm = TRUE)+ 
  labs(title = "Ideological Swing on the Court",
       x= NULL,
       y = "Average Direction")+
  theme_minimal()
```

1 = Conservative

2 = Liberal

Interesting result. This would be the total votes in a year that are considered to be conservative vs. liberal. This is about the political content of the vote, not the partisanship of the justices or the willingness of them to form different coalitions.

```{r}

endpoints <- cases_selected %>%
  group_by(justiceName) %>%
  filter(year == min(year) | year == max(year))

cases_selected |> 
  arrange(year)|>
  group_by(justice)|>
  mutate(justice_avg_direction = mean(direction, na.rm = TRUE))|>
  ggplot(aes(x=year, y = as.factor(justice)))+ 
  geom_line(aes(color = justice_avg_direction),
                size = 2,
            lineend = "round")+
  geom_text(
    data = endpoints,
    aes(label = year),
    vjust = -0.5  # Adjust to move label slightly above the point
  ) +
  scale_y_discrete(
    breaks = cases_selected$justice,
    labels = cases_selected$justiceName
  ) +
  scale_color_gradient2(
    low = "red",    # Color for low values
    mid = "gray",   # Color for midpoint
    high = "darkblue",    # Color for high values
    midpoint = 1.5     # Set the midpoint value for the diverging scale
  ) +
  labs(title = "Ideological Shift and Justice Terms", 
       y = "Justice", 
       x = NULL, 
       legend = "Ideological Average")+
  theme_minimal()
```
