---
title: "supreme_court"
format: html
editor: visual
---

# Supreme Court Polarization

### Goal: To explore whether collaboration is decreasing within the Court due to partisanship. 

```{r}
library(tidyverse)
library(ggiraph)
```

Getting Data from the Supreme Court Database at Washington University Law

```{r}

zip_url <- "http://scdb.wustl.edu/_brickFiles/2024_01/SCDB_2024_01_justiceCentered_Citation.Rdata.zip"

zip_destfile <- "SCDB_2024_01_justiceCentered_Citation.Rdata.zip"

download.file(zip_url, zip_destfile, mode = "wb")

unzip(zip_destfile, exdir = ".")

load("SCDB_2024_01_justiceCentered_Citation.Rdata")

```

I might try this without knowledge of their party affiliations AND independently of it as well.

### Q1: How do I track who votes together? 

What are the groupings? Find votes where they vote the same way, and see how often they appear on the same side together.

### Q2: Is the mean concurring vote total changing over years? 6-3

```{r}
##making a subset
cases_selected <- SCDB_2024_01_justiceCentered_Citation |>
  select(caseId,
         dateDecision,
         majVotes,
         minVotes,
         justice,
         justiceName,
         vote,
         direction,
         majority) |>
  mutate( year = year(dateDecision),
          agreement_percentage = if_else(
            minVotes == 0,
            100,  # Set to 100% if unanimous
            (majVotes / (majVotes + minVotes)) * 100 
            )
          )


```

```{r}
cases_selected |> 
  group_by(year)|>
  summarize(avg_agree_perc = mean(agreement_percentage))|>
  ggplot(aes(x=year, y = avg_agree_perc))+
  geom_point()+
  geom_line()+
  geom_smooth(method = "loess", span = 0.5, 
              method.args = list(degree = 2, #linear or quadratic
                                 family = "symmetric"), #symmetric or gaussian
              na.rm = TRUE)+  
  labs(title = "Supreme Court Agreement Percentage")+
  theme_minimal()

```

Conclusion: There has been a slight upward trajectory to greater agreement in the court. This could just represent partisan sorting or partisan majorities in the modern era. What I need to measure is ideological sorting. Has that gotten worse or better?

### Q3 Does the direction variable actually track changes in the court?

```{r}
cases_selected |>
  group_by(year)|>
  summarize(avg_direction = mean(direction, na.rm = TRUE)) |>
  ggplot(aes(x = year, y = avg_direction)) +
  geom_line()+
  geom_point()+
  geom_smooth(method = "loess", span = 0.5, 
              method.args = list(degree = 2, #linear or quadratic
                                 family = "symmetric"), #symmetric or gaussian
              na.rm = TRUE)+ 
  theme_minimal()
```

1 = Conservative

2 = Liberal
